---
title: "Fish heart activity"
author: Gavin L. Simpson
monobackgroundcolor: "white"
format:
  html:
    toc: true
    html-math-method: katex
    embed-resources: true
    code-line-numbers: false
knitr:
  opts_chunk:
    echo: true
    fig.align: center
highlight-style: github-dark
---

In this activity, you will use R to work with measurements collected during the
fish heart practical. You will first summarise the repeated measurements of your
chosen fish heart from the 2025 class. You will then repeat the calculation for
every recorded fish, summarise the available class data, and visualise the
results.

By the end of the activity, you should be able to:

* select observations using `filter()`,
* calculate a mean, standard deviation, and variance using `summarise()`,
* use `group_by()` to calculate summaries for several fish, and
* create and interpret simple plots using **ggplot2**.

## Download the course files

During the first 15 minutes, create an account at
[Posit Cloud](https://posit.cloud/) and sign in.

Once signed in, locate the **New Project** button in the top right corner, and click on it.
Select the **New Project from Git Repository** option. You will then be prompted to provide the URL of the Git repository. Enter the following url exactly:

> https://github.com/anivet-teaching/dyr-og-vaev-fish-hearts

Now click OK. This will spin up a new RStdudio project in the cloud replete with the files needed for the activity today.

Next, we need some R packages. Locate the the Console; it should be the window on the left of the screen. At the R prompt `>`, paste the line of code below and press <kbd>Enter</kbd> (<kbd>Return</kbd>):
```{r}
#| eval: false
install.packages(c("usethis", "rmarkdown", "readxl", "ggplot2", "dplyr"))
```

This will take a minute or two to download and install the relevant R packages and their dependencies.

## Open the activity and create your script

tLocate the Files pane; it should be in the lower right portion of the screen.
Open the `01-dyr-og-vaev-exercise-1-revised.qmd` file; the file is opened by clicking on its name in the **Files** pane. The file will open in a window above the Console. Look carefully at this window and you'll see a light blue arrow next to the word **Render**. Next this you'll see a small gear or cog icon that looks a bit like this ⚙️. Locate this ⚙️ button and click it. Select the option **Preview in Viewer Pane**. Then click **Render**. The
rendered activity will open in RStudio's Viewer pane which will pop up where the Files pane used to be. Use that page as your
instruction sheet.

Next, create the script in which you will do your analysis:

1. Click the small `+` icon in the top-left corner and select `R Script`, or use
   `File` > `New File` > `R Script`.
2. Save the script as `fish-heart-analysis.R`. The `.R` extension is important.
3. Follow the rendered activity and type the requested code into your script.
4. Run a line or selection from the script with <kbd>Ctrl</kbd> +
   <kbd>Enter</kbd> on Windows or <kbd>Cmd</kbd> + <kbd>Enter</kbd> on macOS.

::: {.callout-important}
The rendered activity is your instruction sheet. Type and run the analysis code
in `fish-heart-analysis.R`; do not edit or run the code chunks in the `.qmd`
file during the activity.
:::

::: {.callout-tip}
### Reuse your code

You do not need to retype code that you have already entered. When a later task
builds on an earlier one, you can copy and paste your existing code within your
script and then change or extend the copied version. This avoids repeatedly
typing the parts of the analysis that stay the same.
:::

## About the data

The data were collected by veterinary medicine students during the 2025 fish
heart practical. Students worked in pairs and repeatedly weighed the same fish
heart. The repeated values therefore describe repeated measurements of one
heart, not different hearts.

Some paper data sheets were unavailable when the results were entered
electronically. Consequently, the electronic dataset does not contain every
fish examined during the practical, and the fish numbers are not consecutive.
Most recorded fish have six measurements, but one has five. This might happen
if a group did not complete the protocol, ran out of time, or if part of its
record was unavailable. We will analyse the measurements that were recorded;
we will not fill in or renumber missing records.

The workbook contains two sheets:

1. `fish-meta`, containing information about the fish; and
2. `fish-weight`, containing the repeated heart-mass measurements.

Today we need only `fish-weight`. We will learn how to join the two sheets in
the follow-up exercise next week.

## Load packages

We need three packages for the analysis. The **readxl** package imports the
Excel data, **dplyr** provides data-wrangling functions, and **ggplot2** creates
plots.

```{r}
#| message: false
library("readxl")
library("dplyr")
library("ggplot2")
```

## Read and inspect the measurements

Use `read_excel()` to import the `fish-weight` sheet. The data workbook is in
the `2025` folder within the downloaded course files.

```{r}
heart_raw <- read_excel(
  "fish-heart-weight.xlsx",
  sheet = "fish-weight"
)

heart_raw
```

This is a data frame: R's version of a rectangular data table. Its columns are
variables and its rows are observations.

Before continuing, discuss these questions with your partner:

1. What does one row represent?
2. Which column identifies the fish?
3. Which column contains the heart-mass measurement?
4. How many rows and columns are in the data?

For today's activity, we only need the fish identifier and heart-mass
measurement. We use `select()` to keep those variables and `rename()` to give
the measurement a more informative name. Heart mass is measured in grams.

```{r}
heart <- heart_raw |>
  select(fish_number, observation) |>
  rename(heart_mass_g = observation)

heart
```

The `|>` symbol is called the **pipe**. When reading the code, think of it as
meaning “and then”. We take `heart_raw`, and then select two variables, and then
rename one of them.

## Choose a fish from the 2025 class

You are **not** selecting the fish that you measured in today's practical.
Instead, imagine that you were a student in the 2025 class and choose one of
the fish recorded by a pair last year.

First, ask R to show the available fish numbers:

```{r}
sort(unique(heart$fish_number))
```

Choose one of the displayed numbers. In the code below, replace `___` with that
number. Do not put quotation marks around the fish number.

```{r}
#| eval: false
chosen_fish <- ___
```

The rendered example uses fish 12 from this point onwards so that you can see
complete example output. In your own script, continue using the fish number you
chose above. Your measurements and summary values may therefore differ from the
ones shown here.

```{r}
#| include: false
chosen_fish <- 12
```

Use `filter()` to keep only measurements belonging to that fish:

```{r}
chosen_measurements <- heart |>
  filter(fish_number == chosen_fish)

chosen_measurements
```

Check the result before calculating a summary:

* Does every row have your chosen fish number?
* How many measurements were recorded?
* Are the measurements identical, or do they vary?

::: {.callout-warning}
If the result contains zero rows, check that `chosen_fish` contains one of the
fish numbers displayed by R. Not every fish's data sheet was available for
electronic entry.
:::

## Summarise your chosen fish

The mean describes the centre of the repeated measurements. The standard
deviation describes how much the measurements vary around their mean. A small
standard deviation means that the repeated measurements were similar; a larger
standard deviation means they differed more.

Variance is another measure of spread. It is the square of the standard
deviation. Standard deviation is measured in grams, whereas variance is
measured in squared grams (g^2^).

```{r}
chosen_summary <- chosen_measurements |>
  summarise(
    n_measurements = n(),
    mean_mass = mean(heart_mass_g),
    sd_mass = sd(heart_mass_g),
    variance_mass = var(heart_mass_g)
  )

chosen_summary
```

Discuss with your partner:

1. What is your estimate of the mass of your chosen fish's heart?
2. What does its standard deviation tell you about your repeated measurements?
3. Why is the variance numerically much smaller than the standard deviation?

## Summarise every recorded fish

We can perform the same calculation separately for every fish. Instead of
filtering to one fish, `group_by(fish_number)` asks R to apply the summary to
each recorded fish number.

::: {.callout-tip}
### Build on your previous summary

You do not need to type the whole summary again. Copy the code you used to
summarise your chosen fish and paste it below that code in your script. Remove
the `filter()` step, replace it with `group_by(fish_number)`, and save the result
as `fish_summary`.
:::

```{r}
fish_summary <- heart |>
  group_by(fish_number) |>
  summarise(
    n_measurements = n(),
    mean_mass = mean(heart_mass_g),
    sd_mass = sd(heart_mass_g),
    variance_mass = var(heart_mass_g)
  )

fish_summary
```

Examine the resulting data frame:

1. What does one row represent now?
2. How many fish are represented in the electronic dataset?
3. Do all fish have the same number of recorded measurements?
4. Which fish has the smallest standard deviation? Which has the largest?

::: {.callout-tip collapse="true"}
### Check

The electronic dataset contains 21 fish. Twenty have six recorded measurements,
and one has five.
:::

## Summarise the available class data

The original rows are repeated measurements, but our biological subject is the
fish. To describe differences among the recorded fish, we therefore summarise
the fish means in `fish_summary`, rather than treating all repeated measurements
as though they came from different fish.

```{r}
class_summary <- fish_summary |>
  summarise(
    n_fish = n(),
    class_mean_mass = mean(mean_mass),
    class_sd_mass = sd(mean_mass),
    class_variance_mass = var(mean_mass)
  )

class_summary
```

The two standard deviations calculated in this activity describe different
kinds of variation:

* A standard deviation in `fish_summary` describes variation among repeated
  measurements of one heart.
* The standard deviation in `class_summary` describes variation in mean heart
  mass among the recorded fish.

Because some data sheets were unavailable, these values describe the available
electronic records rather than every fish examined by the 2025 class.

::: {.callout-tip collapse="true"}
### Check

For the supplied data, the mean of the 21 fish means is approximately 0.4594 g,
and their standard deviation is approximately 0.0855 g.
:::

## Plot the measurements and fish means

We create plots using `ggplot()`. The `aes()` function maps variables in the
data to features of the plot. Here, fish number is mapped to the horizontal
axis and heart mass to the vertical axis.

We convert `fish_number` to a factor only inside the plot so R treats the fish
numbers as labels rather than as positions on a continuous numerical scale.

```{r}
heart |>
  ggplot(
    aes(
      x = factor(fish_number),
      y = heart_mass_g
    )
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    alpha = 0.45
  ) +
  labs(
    x = "Fish number",
    y = "Heart mass [g]"
  ) +
  theme_minimal()
```

`geom_jitter()` moves overlapping points a very small random distance along the
horizontal axis so that we can see the individual measurements more clearly.
The movement is only graphical; it does not change the fish numbers or measured
masses.

Next, add the mean for each fish as a larger red point.

::: {.callout-tip}
### Extend your previous plot

Do not type the entire plot again. In your script, copy the plot code you just
wrote and extend the copied version by adding the new `geom_point()` layer after
`geom_jitter()`. The complete extended plot is shown below.
:::

```{r}
heart |>
  ggplot(
    aes(
      x = factor(fish_number),
      y = heart_mass_g
    )
  ) +
  geom_jitter(
    width = 0.10,
    height = 0,
    alpha = 0.45
  ) +
  # ---- Add this block to show the fish means ----
  geom_point(
    data = fish_summary,
    aes(y = mean_mass),
    colour = "red",
    size = 2.5
  ) +
  # ---- End of the added block ----
  labs(
    x = "Fish number",
    y = "Heart mass [g]"
  ) +
  theme_minimal()
```

Discuss the plot:

1. What do the small grey points represent?
2. What do the larger red points represent?
3. Which fish appears to have the largest mean heart mass? Which the smallest?
4. Which fish appears to have the most variable repeated measurements?
5. Can two fish have similar mean masses but different amounts of variation?

## Plot the distribution of fish means

A histogram shows the distribution of a numerical variable. Here, each
observation in the histogram is the mean heart mass of one recorded fish.

```{r}
fish_summary |>
  ggplot(aes(x = mean_mass)) +
  geom_histogram(
    bins = 8,
    colour = "white"
  ) +
  labs(
    x = "Mean heart mass [g]",
    y = "Number of fish"
  ) +
  theme_minimal()
```

Use the histogram to answer these questions:

1. Around what heart mass are most of the recorded fish located?
2. What are the approximate smallest and largest fish means?
3. Does the distribution appear roughly symmetric, or is one tail longer?
4. Why does the histogram contain 21 values rather than 125?

## Summary

In this activity, you:

1. imported measurements from an Excel workbook;
2. selected and renamed variables;
3. selected a recorded fish from the 2025 class and filtered its measurements;
4. calculated means, standard deviations, variances, and sample sizes;
5. used `group_by()` to repeat a summary for every recorded fish; and
6. visualised both the repeated measurements and the distribution of fish
   means.
